---
title       : "Introduction to Data Visualization"
subtitle    : Some dos and don'ts"
author      : Daniel Anderson
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : zenburn      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
--- 
<style>
em {
  font-style: italic
}
</style>

<style>
strong {
  font-weight: bold;
}
</style>

```{r setup, include = FALSE}
library(knitr)
opts_knit$set(root.dir = "/Users/Daniel/Dropbox/Teaching/ev_base_DM/lectures/intro_data/")
opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

## Remember

* Get R installed: https://cran.r-project.org
* Get RStudio installed: https://www.rstudio.com/products/rstudio/download/

**NOTE:** If you need help with either of the above, please contact me. I'd like everybody to be ready to go **before** we need to use it. Best to get it installed now and make sure it's working so we can troubleshoot if not.

---- .quote

<q> Above all else, show the data

<br>
Edward Tufte

----
## Hypothetical example
* School wants to try out a new reading intervention
* Work with researchers at the UO to design a study
* Kindergarten students who are behind their peers in literacy are selected
* Randomly assign half the students to the intervention, the rest continue with
  "typical" instruction
* Now the study is over - how do we tell if it worked? Visualize it! (and 
  other stuff too)

----
## Barplots
(tried and true)

```{r barplot, echo = FALSE, fig.width = 12, fig.height = 7}
set.seed(110)
d <- data.frame(treatment = rnorm(20, 100, 4),
				control = rnorm(20, 95, 4))
library(tidyverse)
td <- gather(d, condition, score)
# write.csv(td, "/Users/Daniel/Dropbox/Teaching/ev_base_DM/data/plot_data.csv",
#		row.names = FALSE)
td$condition <- as.factor(td$condition)
td$condition_n <- as.numeric(td$condition)

with(td, plot(condition_n, score, 
	type = "n", 
	bty = "n",
	main = "",
	xaxt = "n",
	xlab = "Condition",
	ylab = "Score",
	xlim = c(0.75, 2.25)))
axis(1, 0:3, c("", "Control", "Treatment", ""))
rect(c(0.95, 1.95), 0, c(0.95, 1.95) + .1, rev(apply(d, 2, mean)))
```

-----
## Boxplots
(tried and true)

```{r boxplots, echo = FALSE, fig.width = 12, fig.height = 7}

par(bty = "n")
plot(score ~ condition, data = td)
```

----
## Notched boxplots
(slightly better)

```{r notched_boxplots, echo = FALSE, fig.width = 12, fig.height = 7}
par(bty = "n")
plot(score ~ condition, data = td, notch = TRUE)
```

----
## Stripcharts
Show the data!

```{r stripchart, echo = FALSE, fig.width = 12, fig.height = 7}
par(bty = "n")
plot(score ~ condition_n, data = td, 
	type = "p",
	xaxt = "n",
	xlim = c(0.75, 2.25))
axis(1, 0:3, c("", "Control", "Treatment", ""))
```

----
## Jittered stripcharts
Show the data!

```{r jittered_stripchart, echo = FALSE, fig.width = 12, fig.height = 7}
td$condition_j <- jitter(td$condition_n, amount = .1)
with(td, plot(condition_j, score, 
	type = "p", 
	bty = "n",
	main = "",
	xaxt = "n",
	xlab = "Condition",
	ylab = "Score",
	xlim = c(0.75, 2.25)))
axis(1, at = 0:3, labels = c("", "Control", "Treatment", ""))

```

---
## Combine barplots and jittered stripcharts

```{r barplot_stripchart, echo = FALSE, fig.width = 12, fig.height = 8}
with(td, plot(condition_n, score, 
	type = "n", 
	bty = "n",
	main = "",
	xaxt = "n",
	xlab = "Condition",
	ylab = "Score",
	xlim = c(0.75, 2.25)))
axis(1, 0:3, c("", "Control", "Treatment", ""))
rect(c(0.95, 1.95), 0, c(0.95, 1.95) + .1, rev(apply(d, 2, mean)))
points(td$condition_j, td$score)
```

----
## Combine boxplots and jittered stripcharts

```{r boxplot_jitter, echo = FALSE, fig.width = 12, fig.height = 8}
par(bty = "n")
with(td, plot(condition, score, notch = TRUE))
points(td$condition_j, td$score)
```

----
## Best?

```{r broman_plot, echo = FALSE, fig.width = 12, fig.height = 8}
with(td, plot(condition_j, score, 
	type = "p", 
	bty = "n",
	main = "",
	xaxt = "n",
	xlab = "Condition",
	ylab = "Score",
	xlim = c(0.75, 2.25)))
axis(1, at = 0:3, labels = c("", "Control", "Treatment", ""))


summary <- td %>% 
	group_by(condition) %>% 
	summarize(mean = mean(score),
			  se = sundry::se(score))



low_y_cntrl <- summary$mean[1] - 2*summary$se[1]
low_y_trt <- summary$mean[2] - 2*summary$se[2]
high_y_cntrl <- summary$mean[1] + 2*summary$se[1]
high_y_trt <- summary$mean[2] + 2*summary$se[2]

segments(x0 = c(1.2, 1.8), x1 = c(1.2, 1.8), 
		 y0 = c(low_y_cntrl, low_y_trt),
		 y1 = c(high_y_cntrl, high_y_trt), 
		 lwd = 2)
segments(x0 = rep(c(1.2, 1.8), each = 2) - .025,
		 x1 = rep(c(1.2, 1.8), each = 2) + .025,
		 y0 = c(low_y_cntrl, high_y_cntrl, low_y_trt, high_y_trt),
		 y1 = c(low_y_cntrl, high_y_cntrl, low_y_trt, high_y_trt),
		 lwd = 2)
points(c(1.2, 1.8), summary$mean, cex = 1.5, pch = 22, bg = "white")
```

----
## Some things to avoid

* 3D plots
* Pie charts
* Dual axes
* Restricted axes
* Unnecessary frills (colors, etc)
	+ Show the data as plainly as possible. Let the data speak!

NOTE: The following 10 slides (and the previous plot) inspired/taken from Karl Broman's presentation on graphs (see [here](https://www.biostat.wisc.edu/~kbroman/presentations/graphs2017.pdf))

--- &twocol
## Examples

*** =left

<div align = "center">
<img src = assets/img/broman_plot.png width = 500 height = 500>
</div>

*** =right

<div align = "center">
<img src = assets/img/bad_data_vis/3d_bar.png width = 500 height = 500>
</div>

--- &twocol
## Examples

*** =left

<div align = "center">
<img src = assets/img/broman_plot.png width = 500 height = 500>
</div>

*** =right

<div align = "center">
<img src = assets/img/bad_data_vis/tube_bar.png width = 500 height = 500>
</div>


--- &twocol
## Examples

*** =left

<div align = "center">
<img src = assets/img/broman_plot.png width = 500 height = 500>
</div>

*** =right

<div align = "center">
<img src = assets/img/bad_data_vis/cone_bar.png width = 500 height = 500>
</div>

--- &twocol
## Examples

*** =left

<div align = "center">
<img src = assets/img/broman_plot.png width = 500 height = 500>
</div>

*** =right

<div align = "center">
<img src = assets/img/bad_data_vis/orbs.png width = 500 height = 500>
</div>


--- &twocol
## Examples

*** =left

<div align = "center">
<img src = assets/img/broman_plot.png width = 500 height = 500>
</div>

*** =right

<div align = "center">
<img src = assets/img/bad_data_vis/pie.png width = 500 height = 500>
</div>

--- &twocol
## Examples

*** =left

<div align = "center">
<img src = assets/img/broman_plot.png width = 500 height = 500>
</div>

*** =right

<div align = "center">
<img src = assets/img/bad_data_vis/exploded_pie.png width = 500 height = 500>
</div>


--- &twocol
## Examples

*** =left

<div align = "center">
<img src = assets/img/broman_plot.png width = 500 height = 500>
</div>

*** =right

<div align = "center">
<img src = assets/img/bad_data_vis/donut.png width = 500 height = 500>
</div>

--- &twocol
## Examples

*** =left

<div align = "center">
<img src = assets/img/broman_plot.png width = 500 height = 500>
</div>

*** =right

<div align = "center">
<img src = assets/img/bad_data_vis/area.png width = 500 height = 500>
</div>

--- &twocol
## Examples

*** =left

<div align = "center">
<img src = assets/img/broman_plot.png width = 500 height = 500>
</div>

*** =right

<div align = "center">
<img src = assets/img/bad_data_vis/double_terrain.png width = 500 height = 500>
</div>

--- &twocol
## Examples

*** =left

<div align = "center">
<img src = assets/img/broman_plot.png width = 500 height = 500>
</div>

*** =right

<div align = "center">
<img src = assets/img/bad_data_vis/terrain.png width = 500 height = 500>
</div>

----
## Some great examples: SEDA
Sean Reardon: https://cepa.stanford.edu/seda/overview

<div align = "center">
<img src = assets/img/seda.png width = 800 height = 600>
</div>

----
## Means by district
<div align = "center">
<img src = assets/img/seda_means.png width = 900 height = 500>
</div>

----
## Average gains by district
<div align = "center">
<img src = assets/img/seda_growth.png width = 900 height = 500>
</div>

----
## Mean scores and SES
<div align = "center">
<img src = assets/img/seda_ses.png width = 900 height = 500>
</div>

----
## Mean scores and SES by Race/Ethniciy
<div align = "center">
<img src = assets/img/seda_ses_race.png width = 900 height = 500>
</div>

---
## Other examples: Visualizing scale
* Space stuff: http://imgur.com/a/lGabv
<br>
* Time: http://www.sciencealert.com/watch-this-3-minute-animation-will-change-your-perception-of-time

----
## Some *ggplot* examples

```{r mpgEx3, echo = FALSE, warning = FALSE, message = FALSE, fig.width = 12, fig.height = 8}
library(ggplot2)
theme_set(theme_classic())
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + geom_smooth()
```

----
## Add an additional aesthetic

```{r mpgEx4, echo = FALSE, fig.width = 12, fig.height = 8}
ggplot(mpg, aes(x = displ, y = hwy, color = class)) + geom_point() 
```

----
## Add smooth line for each class
# Too busy

```{r mpgEx5b, fig.height = 8, fig.width = 12, echo = FALSE, warning = FALSE, message = FALSE}
ggplot(mpg, aes(x = displ, y = hwy, color = class)) + geom_point() +
 geom_smooth()
```

----
## Remove SE

```{r mpgEx6a, echo = FALSE, fig.height = 8, fig.width = 12, warning = FALSE, message = FALSE}
ggplot(mpg, aes(x = displ, y = hwy, color = class)) + geom_point() +
 geom_smooth(se = FALSE)
```

---- .segue
# Some things to avoid

----
## Truncated axes
<div align = "center">
<img src = assets/img/bad_data_vis/truncated_axes.png width = 900 height = 500>
</div>

----
<div align = "center">
<img src = assets/img/bad_data_vis/truncated_axes2.png width = 900 height = 500>
</div>

----
## Dual axes
<div align = "center">
<img src = assets/img/bad_data_vis/dual_axes.png width = 900 height = 500>
</div>

----
## Scaling issues
<div align = "center">
<img src = assets/img/bad_data_vis/area_size.png width = 900 height = 500>
</div>

----
## Poor binning choices
<div align = "center">
<img src = assets/img/bad_data_vis/poor_binning.png width = 900 height = 500>
</div>

----
## Some general advice
* Consider the purpose of the plot. 
	+ Relation? Scatterplots 
	+ Distribution? Histogram or density plot
	+ Trend? Line plot, scatterplot with smoother, etc.
* How many variables? What type?
	+ One continuous variable: histogram, density plot, or similar
	+ Two continuous: Scatterplot (if you have lots of data, consider binning)
	+ One categorical one continuous: boxplots, violin plots, bar plots
	+ Two categorical variable? Mosaic plot

----
## One continuous variable
# Histogram
```{r hist, echo = FALSE, fig.height = 7, fig.width = 12, message = FALSE}
ggplot(mpg, aes(hwy)) + geom_histogram()
```

----
## One continuous variable
# Density plot
```{r density, echo = FALSE, fig.height = 7, fig.width = 12, message = FALSE}
ggplot(mpg, aes(hwy)) + geom_density()
```

----
## One continuous variable
# Frequency polygon
```{r freqpoly, echo = FALSE, fig.height = 7, fig.width = 12, message = FALSE}
ggplot(mpg, aes(hwy)) + geom_freqpoly()
```

----
## Consider overlays
```{r overlay, echo = FALSE, fig.height = 8, fig.width = 12, message = FALSE}
ggplot(mpg, aes(hwy)) + geom_histogram(aes(y = ..density..)) + geom_density(col = "blue")
```

----
## Two continuous variables
# Scatterplot

```{r scatter, echo = FALSE, fig.height = 7, fig.width = 12}
ggplot(diamonds, aes(carat, price)) + geom_point()
```

----
## Trend
# Line plot (often with date or time on x-axis)
```{r trend, echo = FALSE, fig.height = 7, fig.width = 12}
ggplot(economics, aes(date, unemploy)) + geom_line()
```

----
## Trend w/smoother

```{r trend_smooth, echo = FALSE, fig.height = 7, fig.width = 12, message = FALSE}
ggplot(economics, aes(date, unemploy)) + geom_line() + geom_smooth()
```

---
## Categorical & Continuous
# Violin plots

```{r violin_plot, echo = FALSE, fig.height = 7, fig.width = 12}
ggplot(mpg, aes(class, hwy)) + geom_violin()
```

---
## Overlay data

```{r violin_plot_data, echo = FALSE, fig.height = 8, fig.width = 12}
ggplot(mpg, aes(class, hwy)) + geom_violin() + geom_jitter(height = 0, width = 0.2)
```

---
## Two categorical variables
# Mosaic plot
```{r mosaic_plot, echo = FALSE, fig.height = 7, fig.width = 12}
mosaicplot(UCBAdmissions, sort = 3:1,
      col = hcl(c(120, 10)),
      main = "Student admissions at UC Berkeley")
```

----
## Don't end up in a blog for wrong reasons
* https://flowingdata.com/2010/05/14/wait-something-isnt-right-here/
* https://flowingdata.com/2009/11/26/fox-news-makes-the-best-pie-chart-ever/

----
## Conclusions
* Essentially never
	+ Use pie charts (use bar charts instead)
	+ Use dual axes (produce separate plots instead)
	+ Truncate axes
	+ Use 3D unnecessarily
	+ Add color for color's sake (this isn't sales)
* Do
	+ Show the data
	+ Be as clear as possible
	+ Let the data tell the story

----
## Next time
Apply what we've talked about today with R and ggplot!

<div align = "center">
<img src = assets/img/ggplot_cheatsheet.png width = 900 height = 500>
</div>


